---
title: "Time series"
output: html_document
date: "2023-06-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(haven)
```


##Weighted moving average
```{r}
weighted_moving_average <- function(x){
  n <- length(x)
  if(n<3){
    print("order must be at least 3")
  } else if(n %% 2 == 0){
    print("order must be odd")
  }  else {
    weights <- c(0.5, rep(1, n-2), 0.5)/(n-1)
    sum(x*weights)
  }
}
```
q <- (n-1)/2

##Time series analysis
```{r}
airline <- read_sav("data/Book 2/airline.sav")
```


###Decomposition
```{r}
pseudo_month <- pseudo %>% count(month_floor) %>% 
  mutate(year = year(month_floor),
         month = month(month_floor),
         n_ave = rollmean(n, k = 3, fill = NA, align = "center"),
         #n_ave_log = log(n_ave),
         #n_ave_sqrt = sqrt(n_ave)
         m_hat = rollapply(n, width = 13, weighted_moving_average, align = "center", fill = NA),
         Yt = n - m_hat) %>% # Yt = st + 'Wt
  filter(year<max(year)) %>% # want complete years: partial years will have strangely high or low Fbar
  group_by(month) %>%
  mutate(Fj = mean(Yt, na.rm = TRUE)) %>% # Raw seasonal factor#
  ungroup() %>%
  group_by(year) %>% # Fbar
  mutate(F_bar = mean(Fj)) %>% # Mean raw seasonal factor
  ungroup() %>%
  mutate(s_hat = Fj - F_bar, # seasonal factor
         Zt = n - s_hat, # Seasonally adjusted series
         Zt_ave = rollmean(Zt, k = 6, fill = NA, align = "center"),
         W_hat = Zt - m_hat, # irregular component
         Xt = m_hat + s_hat + W_hat) # original time series


ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = n))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = n_ave))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = m_hat))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = Yt))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = Fj))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = F_bar))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = s_hat))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = Zt))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = Zt_ave))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = W_hat))
ggplot(pseudo_month) + geom_line(aes(x = month_floor, y = Xt))

sums <- pseudo_month %>% group_by(year) %>% summarise(sum_Fj = sum(Fj), sum_s_hat = sum(s_hat))

```

###Seasonal pattern
```{r}
ggplot(pseudo_month) + geom_line(aes(x = month, y = n, col = year, group = year))
ggplot(pseudo_month) + geom_line(aes(x = month, y = s_hat, col = year, group = year))
```
Lowest April
Highest October

###Exponential smoothing

```{r}

alpha <- 0.9612
observed <- pseudo_month$n
expected <- observed[1]
for(i in 1:(length(observed)-1)){
  expected[i+1] <- alpha*observed[i]+(1-alpha)*expected[i]
}

pseudo_month_exp <- pseudo_month %>% mutate(exp = expected) %>% select(month_floor, n, exp) %>% mutate(err = n-exp, err_squ = err^2)
sse <-  sum(pseudo_month_exp$err_squ)
sse

ggplot(pseudo_month_exp %>% pivot_longer(cols = n:exp, names_to = "names", values_to = "values")) + geom_line(aes(x = month_floor, y = values, col = names))
```


alpha = 0.9612
sse = 10,773,677

alpha = 0.8470155
sse = 10,929,568


```{r}
alpha_var <- seq(0,1,0.0001)
observed <- pseudo_month$n
sse <- vector("integer", 0)

for(j in seq_along(alpha_var)){
  alpha <- alpha_var[j]
  expected <- observed[1]
  for(i in 1:(length(observed)-1)){
    expected[i+1] <- alpha*observed[i]+(1-alpha)*expected[i]
  }
  sse[j] <- sum((observed-expected)^2)
}
min(sse)
alpha_var[sse == min(sse)]

```


